Characterizing edges in signed and vector-valued graphs

### 2017-12-19 edge/intro.tex
pass over edge

In this chapter, we extend the problem of predicting edge types beyond the binary case.
Furthermore, we assume that we are provided with node attributes in addition to the mere graph
topology we used in the previous chapters. Our motivation is twofold: intrinsic and extrinsic.
First, we want an interpretable model that, given two connected nodes, can explain why they are
linked. This lead us to use a linear model, for it is both easily amenable to
interpretation and computationally efficient. Second, edge-labeled graphs are susceptible to further
studies by existing networks analysis methods, as we will describe shortly. Thus, our method is a
useful way of enriching data to extract new insights from it.

As demonstrated by our prior extensive coverage, predicting edge signs is an interesting and
challenging problem in its own right. Yet it is natural to extend and generalize it.  First, there
are other edge binary classification problems that arise in graphs. For instance, predicting whether
a link in a social network is strong or weak~\autocites{communityWeakTies14}{communityWeakTies17};
predicting whether two authors in a citation network are advisor-advisee or regular
coauthors~\autocite{Advisor10}; predicting whether two persons in a company network have a
relationship of manager-subordinate or not~\autocite{Manager07}; predicting whether two proteins
interact physically or genetically; or whether two neurons have a chemical or ionic link between
them~\autocite{BioMultiplex15}.

@@

\medskip
Besides predicting potentially more than two classes, the other difference with the setting of the
previous chapters is that we now assume that nodes are associated with attributes, which we call
\emph{profiles}. One could argue this makes our framework less general, for attributes are not
always available, or may come with missing values. Another recent finding is that, despite a natural
assumption, the nodes profiles in attributed graphs are not necessarily correlated with known ground
truth communities~\autocite{FortunatoGroundTruth14}, casting further doubt on the usefulness of that
extra information.
This raises the question of why we cannot simply extend the troll/trust method of counting the type
of edges incident to every node and learn a model from such features.
A similar tension between relying solely on always available yet sparse topological information or
leveraging possibly richer attributes information also exists in link
prediction~\autocite{linkPredSurvey16}. %\stodo{What they actually write is \enquote{Topology-based
%techniques are more versatile than attribute-based methods since they are not domain specific} which
%sounds a bit unfair: only the attributes are domain specific, not necessarily the methods
%themselves}.
% examples of using node attributes for link prediction, taken from that survey:
% https://doi.org/10.1007/s11432-014-5237-y
% - http://dl.acm.org/citation.cfm?id=2124378
% - using graph kernel https://doi.org/10.1016/j.dss.2012.09.019
% - matrix factorization “these latent features (learned from topology) may be combined with
%   optional explicit features for nodes or edges, which yields better performance than using either
%   type of feature exclusively.” https://doi.org/10.1007/978-3-642-23783-6_28
However, the presence of profiles is motivated by the last fundamental difference with the previous
problems. In this chapter, we do not have access to any direct supervision in the form of labels.
Rather than a strict classification problem, we are thus faced with unsupervised clustering, guided
by the information contained in node profiles. Note that contrary to the relation between \pcc{} and
\esp{}, this time the objects we seek to cluster are edges and not nodes. Furthermore, we stress
that the profiles crucially increases the interpretability of our models.

>>>>

In this chapter, we extend the problem of predicting edge types beyond the binary case.
Furthermore, we assume that we are provided with node attributes in addition to the mere graph
topology we used in the previous chapters. Our motivation is twofold: intrinsic and extrinsic.
First, we want an interpretable model that, given two connected nodes, can explain why they are
linked. This leads us to use a linear model, for it is both easily amenable to
interpretation and computationally efficient. Second, edge-labeled graphs are susceptible to further
studies by existing networks analysis methods, as we will describe shortly. Thus, our method is a
useful way of enriching data to extract new insights from it.

As demonstrated by our prior extensive coverage, predicting edge sign is an interesting and
challenging problem in its own right. Yet it is natural to extend and generalize it.  First, there
are other edge binary classification problems that arise in graphs. For instance, predicting whether
a link in a social network is strong or weak~\autocites{communityWeakTies14}{communityWeakTies17};
predicting whether two authors in a citation network are advisor-advisee or regular
coauthors~\autocite{Advisor10}; predicting whether two persons in a company network have a
relationship of manager-subordinate or not~\autocite{Manager07}; predicting whether two proteins
interact physically or genetically; or whether two neurons have a chemical or ionic link between
them~\autocite{BioMultiplex15}.

@@

\medskip
Besides predicting potentially more than two classes, the other difference with the setting of the
previous chapters is that we now assume that nodes are associated with attributes, which we call
\emph{profiles}. One could argue this makes our framework less general, for attributes are not
always available, or may come with missing values. Another recent finding is that, despite a natural
assumption, the nodes profiles in attributed graphs are not necessarily correlated with known ground
truth communities~\autocite{FortunatoGroundTruth14}, casting further doubt on the usefulness of that
extra information.
This raises the question of why we cannot simply extend the troll/trust method of counting the type
of edges incident to every node and learn a model from such features.
A similar tension between relying solely on always available albeit sparse topological information or
leveraging possibly richer attributes information also exists in link
prediction~\autocite{linkPredSurvey16}. %\stodo{What they actually write is \enquote{Topology-based
%techniques are more versatile than attribute-based methods since they are not domain specific} which
%sounds a bit unfair: only the attributes are domain specific, not necessarily the methods
%themselves}.
% examples of using node attributes for link prediction, taken from that survey:
% https://doi.org/10.1007/s11432-014-5237-y
% - http://dl.acm.org/citation.cfm?id=2124378
% - using graph kernel https://doi.org/10.1016/j.dss.2012.09.019
% - matrix factorization “these latent features (learned from topology) may be combined with
%   optional explicit features for nodes or edges, which yields better performance than using either
%   type of feature exclusively.” https://doi.org/10.1007/978-3-642-23783-6_28
However, the presence of profiles is motivated by the last fundamental difference with the previous
problems. In this chapter, we do not have access to any direct supervision in the form of labels.
Rather than a strict classification problem, we are thus faced with unsupervised clustering, guided
by the information contained in node profiles. Note that contrary to the relation between \pcc{} and
\esp{}, this time the objects we seek to cluster are edges and not nodes. Furthermore, we stress
that the profiles crucially increase the interpretability of our models.
